The Community Land Model, version 5: \\ Parameter Perturbation Experiment

### 2023-04-26 ms/ms.tex
update
restore ms

Water availability, land temperature extremes, fire risk, crop productivity, and the strength of the terrestrial carbon sink.
Our certainty in model projections varies by domain and generally decreases as the time horizon is extended.
Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations (Fried, Arora). 
A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate (Bonan and Doney?).
But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns we should expect predictions to become more skillful over time.

While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations.
Within the model development process, there is a larger emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
At this point, complexity of land models has increased to the point

Probabilistic forecasts

Models have parameters.
Some can be measured, but many cannot.

Parametric uncertainty vs. structural uncertainty

\section{Experiment Description}
\label{methods}
\subsection{Model description}
\label{sect:md}
The Community Terrestrial System Model (CTSM) is developed by the CESM Land Model Working Group (LMWG) and maintained at the National Center for Atmospheric Research (NCAR). This experiment utilizes the Community Land Model configuration of CTSM, version 5.1 (CLM5.1). The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.

@@

We used k-means clustering to identify groups of grid cells with similar climatologies based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase. We selected one representative grid cell from each cluster, whichever is located nearest the cluster centroid in climate space. The set of representative grid cells comprise a `sparsegrid', which are used in lieu of a coarse grid. A paint-by-number approach is used to recompose mapped output and compute global means, where the output from the representative grid cell is substituted for all members of the cluster cohort.

Clustering was based on a subset of 18 meaningful CLM variables (Table \ref{tab:sg}). The clustering algorithm received 12 observations of each variable per grid cell, namely the mean and interannual variability computed for six 30-year climatology windows (1865-1894, 1895-1924, ... , 1985-2014). Clusters were delineated to equalize the multi-dimensional variance across the user-specified number of groups, $k$. We tested 15 values of $k$, ranging from 10 to 800. Based on ILAMB2.5 benchmarking \cite{collier2018} against the full grid output, we opted for a 400-cluster sparsegrid, to balance computational cost against model fidelity (Supp Figure \ref{supp:ilamb}). Because our emphasis is on vegetated regions, we masked out Antarctica within the clustering algorithm, whereby we do not provide any output below 60$^\circ$S.

@@

\subsection{Experimental Design}
In the preparation for this experiment, 206 CLM-BGC parameters were identified. We decided to vary each parameter once at a time, to a minimum and maximum values. To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter. In some cases literature values were available, but in the vast majority expert judgment was used. The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.

Each simulation lasted 150 years, with the first 140 for spin-up, followed by a 10-year period for analysis purposes. We opted for six different forcing scenarios to understand the intersection of parameter effects and climate change (Table \ref{tab:exps}). The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and is the default forcing data for CLM5 \cite{lawrence2019}. We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings. Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the mean climate change signal. We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.

@@

\section{Results}

\begin{figure}[h]
\centering
\includegraphics[width=30pc]{../figs/sims.pdf}
\caption{The approximate number of simulations afforded by 1 million core-hours for a range of CLM configurations. Configurations are labeled according to spin-up procedure and resolution, with `sg' signifying sparsegrid. The inset map shows the locations of the 400 sparse grid cells as the red dots. See Section \ref{methods} for spin-up and sparsegrid details. My speed calculations here are a bit janky, assume linear scaling of cost for increasing time or space (zqz). }
\label{fig:sims}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=40pc]{../figs/sparsegrid_gpp.pdf}
\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation. We opted for 400 clusters to balance computational cost against representativeness. For reference there are 22648, 5666, and 1764 land gridcells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively.}
\label{}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=35pc]{../figs/violins.pdf}
\caption{Distributions of global annual gross primary production (GPP) and net biome production (NBP) across the six forcing scenarios. See Section \ref{sect:exps} for scenario descriptions. The distributions are concentrated around the default parameterization with long tails, indicating that a small proportion of parameters have large effects on these two variables. In the case of NBP, parameter effects appear larger than scenario effects.}
\label{}
\end{figure}

@@

\section{Discussion}
\section{Conclusion}
\section*{Open Research Section}
This section MUST contain a statement that describes where the data supporting the conclusions can be obtained. Data cannot be listed as ''Available from authors'' or stored solely in supporting information. Citations to archived data should be included in your reference list. Wiley will publish it as a separate section on the paper’s page. Examples and complete information are here:
https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors

>>>>

Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many variables represented within the land component of Earth System Models.
Understanding how such states respond to carbon dioxide concentration, and how we expect carbon dioxide concentration to evolve over time, is at the center of climate change research.  
Our certainty in climate model projections varies by domain and generally decreases with extended time horizons.
Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. 
A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{bonan2018}.
But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, we would expect predictions to become more skillful over time.

While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations.
Within the day-to-day model development process, there is a greater emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
However, the complexity of land models has increased to the point where understanding the full span of model function has become challenging \cite{fisher2020}, and manual model calibration is becoming intractable \cite{dagon2020}.
Maintaining, improving, and interpreting complex land models benefits from thoughtful investment in software to automate and routinize important components of the development process, e.g. \citeA{collier2018}.

One such fundamental model development step is parameter sensitivity testing and/or parameter optimization.
Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017}.
When applied systematically, the result is a parameter perturbation ensemble (PPE), alternatively termed a perturbed physics ensemble.
PPEs have played an important role in assessing projection uncertainty in climate models.
They also form a basis for automated model calibration.
Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes. 
In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
In this paper we present that dataset, how it was produced, and its potential applications.

We are working with the Community Land Model (CLM), the land component of the Community Earth System Model, which is developed and maintained at the National Center for Atmospheric Research (NCAR).
This work follows up on \citeA{dagon2020}, which began the process of systematically varying CLM parameters in service of automated model calibration.
We extend that work by sampling a more exhaustive set of parameters across a broader range of land model processes.
A set of 206 parameters were varied independently (one-at-a-time) across a range of climate scenarios.
This dataset has already demonstrated great utility for diagnosing parameter effects, while the software and modeling infrastructure has greatly our expanded our capability to generate insights about CLM dynamics and uncertainties.

\section{Experiment Description}
\label{methods}
\subsection{Model description}
\label{sect:md}
The Community Terrestrial System Model (CTSM) is developed by the CESM Land Model Working Group and maintained at the National Center for Atmospheric Research (NCAR). This experiment utilizes the Community Land Model configuration of CTSM, version 5.1 (CLM5.1). The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.

@@

We used k-means clustering to identify groups of grid cells with similar climates based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase. We selected one representative grid cell from each cluster to stand in for the entire cluster. The representative grid cell is whichever is located nearest the cluster centroid in climate space. The set of representative grid cells comprise a `sparsegrid', which are used in lieu of a `coarse' grid. A paint-by-number approach is used to recompose mapped output and compute global means, where the output from the representative grid cell is substituted for all members of the cluster cohort.

Clustering was based on a subset of 18 meaningful CLM variables (Table \ref{tab:sg}). The clustering algorithm analyzed 12 observations of each variable per grid cell, namely the mean and interannual variability computed for six 30-year climatology windows (1865-1894, 1895-1924, ... , 1985-2014). Clusters were delineated to equalize the multi-dimensional variance across the user-specified number of groups, $k$. We tested 15 values of $k$, ranging from 10 to 800. Based on ILAMB2.5 benchmarking \cite{collier2018} against the full grid output, we opted for a 400-cluster sparsegrid, to balance computational cost against model fidelity (Supp Figure \ref{supp:ilamb}). Because our emphasis is on vegetated regions, we masked out Antarctica within the clustering algorithm, whereby we do not provide any output below 60$^\circ$S.

@@

\subsection{Experimental Design}
In the preparation for this experiment, 206 CLM-BGC parameters were identified. We decided to vary each parameter independently, to a low and high value. To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter. In some cases literature values were directly utilized, but in the majority of cases, expert judgment was used. The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.

Each simulation comprised 150 years, with the first 140 for spin-up, followed by a 10-year period for analysis. We opted for six different forcing scenarios to understand the intersection of parameter effects and climate change (Table \ref{tab:exps}). The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and is the default forcing data for CLM5 \cite{lawrence2019}. We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings. Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to a derived climate change signal. We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.

@@

\section{Results}
A major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times faster than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
There are 22648, 5666, and 1764 land gridcells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibration criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.

\begin{figure}[h]
\centering
\includegraphics[width=30pc]{../figs/sims.pdf}
\caption{The approximate number of simulations afforded by 1 million core-hours for a range of CLM configurations. Configurations are labeled according to spin-up procedure and resolution, with `sg' signifying sparsegrid. The inset map shows the locations of the 400 sparse grid cells. See Section \ref{methods} for spin-up and sparsegrid details. (My speed calculations here are a bit janky, assume linear scaling of cost for increasing time or space zqz) }
\label{fig:sims}
\end{figure}

Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
stronger the relationship between output from the full grid and the sparsegrid. Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sggpp}). We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
\begin{figure}[h]
\centering
\includegraphics[width=40pc]{../figs/sparsegrid_gpp.pdf}
\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation. We opted for 400 clusters to balance computational cost against representativeness.}
\label{fig:sggpp}
\end{figure}

Parameters were varied across a wide variety of CLM subcomponents (Figure \ref{fig:params}). A total of 206 parameters were targeted for sensitivity testing. Many parameters were hard-coded, but have now been extracted to the CLM parameter file for easier manipulation. In some cases new parameters were introduced to allow perturbation of important processes. For example a sand perturbation factor ($sand\_pf$) was introduced in order to perturb the percent sand in the soil column without creating a new surface dataset. A spreadsheet detailing all of the parameters, and their ranges, is provided in the supplementary material and available online (zqz link).

@@

\begin{figure}[h]
\centering
\includegraphics[width=35pc]{../figs/violins.pdf}
\caption{Distributions of global annual gross primary production (GPP) and net biome production (NBP) across the six forcing scenarios. See Section \ref{sect:exps} for scenario descriptions. The distributions are concentrated around the default parameterization with long tails, indicating that a small proportion of parameters have large effects. Especially in the case of NBP, parameter effects can be larger than scenario effects.}
\label{}
\end{figure}

@@

\section{Discussion}

Major discussion points:
\begin{itemize}
\item need for speed
\item parameter effects can be quite large
\item parameter priors are difficult to establish
\item mean and climate response variables may vary
\item community resource
\item invest in infrastructure
\end{itemize}

\section{Conclusion}
\section*{Open Research Section}
This section MUST contain a statement that describes where the data supporting the conclusions can be obtained. Data cannot be listed as ''Available from authors'' or stored solely in supporting information. Citations to archived data should be included in your reference list. Wiley will publish it as a separate section on the paper’s page. Examples and complete information are here:
https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
